---
title: "Getting Started with NearRepeatLite"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with NearRepeatLite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## What is NearRepeatLite?

`NearRepeatLite` is a memory-efficient implementation of the Knox test for
near-repeat spatio-temporal clustering. It is a drop-in replacement for
`NearRepeat()` from [Steenbeck's NearRepeat package](https://github.com/wsteenbeek/NearRepeat),
designed for use when large datasets cause memory allocation errors.

The standard `NearRepeat()` function calls `stats::dist()` to compute pairwise
distances between all events, producing a condensed matrix of *n(n−1)/2* values
held in RAM simultaneously. For a dataset of ~79,000 events this requires
approximately 50 GB — infeasible on standard hardware.

`NearRepeatLite()` processes one row of the distance matrix at a time. Only
the observed count matrix and within-range pair indices are retained, reducing
peak RAM from O(n²) to O(n + within-range pairs). For the Oslo dataset, peak
usage drops from ~50 GB to under 1 GB.

| n events | NearRepeat() peak | NearRepeatLite() peak |
|---|---|---|
| 5,000 | ~200 MB | ~50 MB |
| 10,000 | ~800 MB | ~100 MB |
| 50,000 | ~20 GB | ~400 MB |
| 79,294 | ~50 GB | < 1 GB |

Results are identical in observed counts (deterministic). Knox ratios and
p-values differ only due to Monte Carlo noise, which is expected.

## Installation

```r
# Install remotes if needed
install.packages("remotes")

# Install NearRepeatLite from GitHub
remotes::install_github("aallvin/NearRepeatLite")
```

## Quick start

```{r quickstart}
library(NearRepeatLite)

# Load the synthetic Oslo example dataset (500 events, 2015-2019)
data(oslo_sim)
head(oslo_sim)
```

## Defining bands

Spatial and temporal bands are defined as numeric breakpoint vectors.
Intervals are left-closed and right-open by default, e.g. `[0, 1)`, `[1, 101)`.
Breaks are chosen so that integer Manhattan distances fall in the intended band:
a break at 101 (not 100) ensures distance = 100 m falls inside the "1–100 m"
band rather than the next one.

```{r bands}
# Spatial: repeat location (0 m) + 100 m intervals to 500 m
sds <- c(0, 1, 101, 201, 301, 401, 501)

# Temporal: same day + four weekly bands
tds <- c(0, 1, 8, 15, 22, 29)
```

## Running the Knox test

Set a seed before calling `NearRepeatLite()` for reproducible Knox ratios
and p-values. Use `nrep = 99` for exploratory runs and `nrep = 999` for
publication-ready results.

```{r run, cache = TRUE}
set.seed(1519)
result <- NearRepeatLite(
  x    = oslo_sim$x,
  y    = oslo_sim$y,
  time = oslo_sim$date,
  sds  = sds,
  tds  = tds,
  nrep = 99
)
```

## Inspecting results

The function returns a `"knox"` class object with four elements:

```{r results}
# Observed number of crime-pair counts per band combination
result$observed

# Knox ratios (observed / mean simulated)
round(result$knox_ratio, 2)

# Monte Carlo p-values
round(result$pvalues, 3)
```

A Knox ratio > 1 means more pairs were observed in that band combination than
expected by chance. Following Ratcliffe (2008), cells with Knox ratio ≥ 1.2
and p < 0.05 are considered practically significant near-repeat patterns.

## Visualising results

The output is fully compatible with `plot.knox()` from Steenbeck's NearRepeat
package:

```r
# Install NearRepeat for plotting (not required to run NearRepeatLite)
# remotes::install_github("wsteenbeek/NearRepeat")
# library(NearRepeat)
# plot(result)
```

## When to use NearRepeatLite vs. NearRepeat

| Situation | Recommendation |
|---|---|
| n < ~15,000 events | Either; `NearRepeat()` is faster for small datasets |
| n ≥ ~15,000, or R reports allocation errors | `NearRepeatLite()` |
| Replicating Steenbeck's published examples | `NearRepeat()` |
| Pooled multi-year analysis of large crime types | `NearRepeatLite()` |

## References

Ratcliffe, J. H. (2008). *Intelligence-Led Policing*. Cullompton: Willan.

Steenbeck, W. (2018). *NearRepeat: Near Repeat Calculator in R.*
R package. <https://github.com/wsteenbeek/NearRepeat>
